---
title: "Homework 3"
author: ""  # Your name here
date: "`r Sys.time()`"
output: html_document
---

```{r setup, message=FALSE}
library(tidyverse)  # loads many useful packages

set.seed(1)  # Ensures reproducable results

theme_set(theme_minimal())  # My personal preference for less ugly default plots
```

```{r 1.1}
# Import your cleaned data and do any necessary summarization


```

# Part 1: P-values

```{r 1.1.a}
# Perform a t-test for no reward forgettable vs no reward memorable using t.test()


```


```{r 1.1.b}
# Perform a t-test for low reward forgettable vs low reward memorable


```

```{r 1.1.c}
# Perform a t-test for high reward forgettable vs high reward memorable


```

```{r 1.1.d}
# Perform a t-test for no reward forgettable vs low reward forgettable


```

```{r 1.1.e}
# Perform a t-test for low reward forgettable vs high reward forgettable


```

```{r 1.1.f}
# Perform a t-test for no reward forgettable vs high reward forgettable


```

```{r 1.1.g}
# Perform a t-test for no reward memorable vs low reward memorable


```

```{r 1.1.h}
# Perform a t-test for low reward memorable vs high reward memorable


```

```{r 1.1.i}
# Perform a t-test for no reward memorable vs high reward memorable


```

For any 3 of the above tests, discuss the null and alternative hypotheses. Then write the results as they would be presented in a research paper.






Discuss generally the findings from the above t-tests, what you can conclude, and what if any caveats exist.






# Part 2: Multiple Comparisons Correction

```{r 2.4}
# Create a function that takes in a vector of unordered p-values and
# returns a bool for each p-value indicating if the null is rejected
# Assume the original alpha = 0.05
# Show your function works on a toy set of p-values
# ?length

bonferroni = function(p) {
  
}
```

```{r 2.4.a}
# Use your bonferroni function to determine which p-values from part 1 survive correction


```

Report the new alpha value given a Bonferroni correction on the tests used above and describe which, if any, tests survive the correction.





Rewrite the results for the tests taking into account the correction.





```{r 2.5}
# Create a function that takes in a vector of unordered p-values and
# returns a bool for each p-value indicating if the null is rejected
# Show your function works on a toy set of p-values
# ?sort

fdr_correction = function(p) {
  
}
```

```{r 2.5.a}
# Use your function to correct the tests performed in part 1



```

Describe the new p-value threshold and which tests, if any, survive correction.




Write up the results of the tests.





```{r 2.5.c}
# Perform an fdr correction using p.adjust()



```
