---
title: "Homework 3"
author: ""  # Your name here
date: "`r Sys.time()`"
output: html_document
---

```{r setup, message=FALSE}
library(tidyverse)  # loads many useful packages

set.seed(1)  # Ensures reproducable results

theme_set(theme_minimal())  # My personal preference for less ugly default plots
```

```{r 1.1}
# Import your cleaned data and do any necessary summarization


```

# Part 1: P-values

```{r 1.1.a}
# Perform a t-test for no reward forgettable vs no reward memorable using t.test()


```


```{r 1.1.b}
# Perform a t-test for low reward forgettable vs low reward memorable


```

```{r 1.1.c}
# Perform a t-test for high reward forgettable vs high reward memorable


```

```{r 1.1.d}
# Perform a t-test for no reward forgettable vs low reward forgettable


```

```{r 1.1.e}
# Perform a t-test for low reward forgettable vs high reward forgettable


```

```{r 1.1.f}
# Perform a t-test for no reward forgettable vs high reward forgettable


```

```{r 1.1.g}
# Perform a t-test for no reward memorable vs low reward memorable


```

```{r 1.1.h}
# Perform a t-test for low reward memorable vs high reward memorable


```

```{r 1.1.i}
# Perform a t-test for no reward memorable vs high reward memorable


```

For any 3 of the above tests, discuss the null and alternative hypotheses. Then write the results as they would be presented in a research paper.






Discuss generally the findings from the above t-tests, what you can conclude, and what if any caveats exist.






# Part 2: Multiple Comparisons Correction

```{r 2.4}
# Conduct Bonferroni correction from scratch
# Put all your t-test p-values into a dataframe and include a column with their original significance (TRUE/FALSE)

```

```{r 2.4.a}
# Write code to determine which tests survive Bonferroni correction by adding a new column to your dataframe (TRUE/FALSE)
# Verify that the columns important for answering this question can be seen on your knitted html document


```

Report the new alpha value given a Bonferroni correction on the tests used above and describe which, if any, tests survive the correction.





Rewrite the results for the tests taking into account the correction.





```{r 2.5}

# Write code to determine which tests survive Benjamini Hochberg correction (with q=.05) by adding a new column to your dataframe (TRUE/FALSE)
# Verify that the columns important for answering this question can be seen on your knitted html document



```

Describe the new p-value threshold and which tests, if any, survive correction.




Write up the results of the tests.



Is this code flexible to other datasets? If not, what would you need to do to use it with other data?



```{r 2.5.c}
# Perform an fdr correction using p.adjust(). Add columns to your dataframe showing these adjusted p-values and their significance.
# Verify that the columns important for answering this question can be seen on your knitted html document.



```
