---
title: "Homework 2"
author: ""  # Your name here
date: "`r Sys.time()`"
output: html_document
---

```{r setup, message=FALSE}
library(tidyverse)  # loads many useful packages

set.seed(1)  # Ensures reproducable results

theme_set(theme_minimal())  # My personal preference for less ugly default plots
```

# Part 1: Normality and t-tests

```{r 1.1}
# Import your cleaned data


```

```{r 1.2}
# Create an overlayed histogram for memorable and forgettable images
# Use alpha=0.5 and position='identity' to overlay histograms
# https://stackoverflow.com/a/6958052/10067832


```

Discuss the shape of these distributions.



```{r 1.4}
# Perform an independent samples t-test for memorable vs forgettable images from scratch
# Then perform the same test with t.test()


```

```{r 1.5}
# Perform a paired samples t-test for memorable vs forgettable images from scratch
# Then perform the same test with t.test()


```

Discuss which test is best.





```{r 1.6.b}
# Get the p-value for the correct t-test using pt()


```


Write up the results of the t-test as would be appropriate for a scientific paper using inline r code.

Example:

```{r example}
example_p_value = 0.02
```

The p-value is `r example_p_value`.





## Part 2: Effect Size

```{r 2.7}
# Calculate the effect size for your test from scratch


```

Write up your results again, this time including the effect size.



## Part 3: Permutation Tests
 
```{r 3.9.a}
# Create a function that will return a random permutation, to be used in a permutation test


```

```{r 3.9.b}
# Conduct 1000 permutations


```

```{r 3.9.c}
# Plot a histogram of your permutations with the observed difference represented by a vertical line


```

```{r 3.9.d}
# Calculate a p-value for your permutation test


```

Write up your permutation results.





## Part 4: Bootstrapping Confidence Intervals

```{r 4.10}
# Calculate the confidence interval around the mean for memorable and forgettable images


```

```{r 4.11.a}
# Create a function that will generate a bootstraped sample
# and return the sample mean
# ?sample_frac

get_sample = function(data) {
  
}
```

```{r 4.11.b}
# Use replicate to generate 1000 sample means for memorable images


```

```{r 4.11.c}
# Use replicate to generate 1000 sample means for forgettable images


```

```{r 4.11.d}
# Calculate the 95% CIs for memorable and forgettable images


```

Compare your parametric CIs to your bootstrapped CIs.




